--- 
title: "Single Cell DNA Analysis Lab"
author: "Andrew McPherson"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is the lab practical for a module on single cell DNA analysis for the CBW Cancer Analysis Workshop"
---

# Lab Overview

This book contains lab exercises for Module 05: Single Cell Genomics - DNA, as part of the Canadian Bioinformatics Workshop Cancer Analysis (CAN) Workshop.

## Instructions

* Labs will use **Rstudio**.
* Each lab will be provided as blocks of R code you will copy-paste into Rstudio to follow the building of the predictor. Code blocks to be pasted will be in yellow like so.

```{r,class.source="codeblock",eval=FALSE}
x <- 5
```
If you hover your mouse over the block, you should see a little clipboard icon. Clicking on that icon will allow you to copy the block in one shot. I would use that.

In each case, the result of executing a code block is shown directly after the code, in grey blocks like the one below. These are not intended to be pasted in. 

```{r,eval=FALSE}
5
```